---
title: "Final_CommonGarden_Markdown"
date: "2025-04-01"
output: html_document
---


baseline data setup 
```{r}
library(dada2) #call on dada2 package
library(tidyverse) # call on package for data manipulating
library(phyloseq) #package required for decontamination, also good for many analyses
library(decontam) #'decontam

setwd("~/Desktop/Aim 1+2/aims_1_2_seq") #set working directory


path <- "fastq-primers-trimmed" #setting the relative path to the trimmed fastq files
fns <- list.files(path, pattern=".fastq", full.names=TRUE) #defining the file names
head(fns) #checking that I properly navigated to the correct path

#now define the forward and reverse files
fnFs <- list.files(path, pattern="_R1_001.fastq", full.names=TRUE)
fnRs <- list.files(path, pattern="_R2_001.fastq", full.names=TRUE)
head(fnFs)

q_plot_F<-plotQualityProfile(fnFs[1:2]) #plot quality of forward reads, choosing first two samples and negatives

pdf(paste("/Users/russ/Desktop/q_plot_F.pdf"), width = 10, height = 10)
print(q_plot_F)
dev.off()

q_plot_R <- plotQualityProfile(fnRs[1:2]) #plot quality of reverse reads, choosing first two samples and negatives

pdf(paste("/Users/russ/Desktop/q_plot_R.pdf"), width = 6, height = 4)
print(q_plot_R)
dev.off()

#dereplication

#filtering with default paramaters and truncating to length based on quality profiles
filtFs <- file.path(path, "filtered", basename(fnFs)) #directory for filtered forward reads
filtRs <- file.path(path, "filtered", basename(fnRs)) #directory for filtered reverse reads

out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(275, 225)) #fix truncLen based on quality score stuff

#denoising and removing chimeras
errF <- learnErrors(filtFs, multi=TRUE)
errR <- learnErrors(filtRs, multi=TRUE)

#plotting error models
e_plot_F<-plotErrors(errF, nominalQ=TRUE)
e_plot_R<-plotErrors(errR, nominalQ=TRUE)

pdf(paste("/Users/russ/Desktop/e_plot_F.pdf"), width = 6, height = 4)
print(e_plot_F)
dev.off()

pdf(paste("/Users/russ/Desktop/e_plot_R.pdf"), width = 6, height = 4)
print(e_plot_R)
dev.off()

#DADA 
ddF <- dada(filtFs, errF, pool="pseudo", multi=TRUE) #note the pseudopooling option and the detection of singletons is enabled
ddR <- dada(filtRs, errR, pool="pseudo", multi=TRUE)

merged <- mergePairs(ddF, filtFs, ddR, filtRs, verbose=TRUE, justConcatenate = FALSE) #concatenate (meshes reads together).  False bc i don't want concatenated. merge them bc there is overlap only not merge with miniseq bc no overlap

#construct the table
table.chimeras <- makeSequenceTable(merged)
dim(table.chimeras)

#remove chimeras
table.no.chimeras <- removeBimeraDenovo(table.chimeras, multi=TRUE, verbose=TRUE)
sum(table.no.chimeras)/sum(table.chimeras)

#export this table as a text file
write.table(t(table.no.chimeras), file = 'table.tsv', sep = "\t", row.names = TRUE, col.names=NA, quote=FALSE)

#at this point I'd also like to export the sequences
uniquesToFasta(table.no.chimeras, fout='rep-seqs.fasta', ids = colnames(table.no.chimeras))

#check the pipeline (from primer removal and beyond)
getN <- function(x) sum(getUniques(x))


summary_tab <- data.frame(row.names=fnFs, dada2_input=out[,1],
                          filtered=out[,2], ddF=sapply(ddF, getN),
                          ddR=sapply(ddR, getN), merged=sapply(merged, getN),
                          table.no.chimeras=rowSums(table.no.chimeras),
                          final_perc_reads_retained=round(rowSums(table.no.chimeras)/out[,1]*100, 1))


summary_tab


track <- cbind(out, sapply(ddF, getN), sapply(ddR, getN), sapply(merged, getN), rowSums(table.no.chimeras))
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- basename(fnFs)
head(track)
write.csv(track, "pipeline_stats.csv")

#CLASSIFYING TO ASSIGN TAXONOMY
system.time(tax <- assignTaxonomy(table.no.chimeras, "silva_nr99_v138.1_train_set.fa.gz", 
                                  multi=TRUE))
unname(head(tax))

write.table(tax, file = 'taxonomy.tsv', sep = "\t", row.names = TRUE, col.names=NA, quote=FALSE)

#DECONTAMINATING
#The method I am going to use removes sequences that occur more often in negative controls than they do in real samples.

library(readxl)
#reading in the metadata and creating a phyloseq object
metadata <- read_xlsx("/Users/russ/Desktop/Aim 1+2/aims 1+2 seq data/metadata.xlsx") #importing the metadata
#remove NAs from sequencing facility (positive samples that don't matter, and rename rows)
metadata <- metadata[!is.na(metadata$seq_file), ]
row.names(metadata) <- metadata$seq_file #altering its format so row names are the sample ids


#creating a 'phyloseq' object
trans_table.no.chimeras<-t(table.no.chimeras)
otu<-otu_table(trans_table.no.chimeras, taxa_are_rows = TRUE)
otu<-t(otu)
samp<-sample_data(metadata)
row.names(samp) <- samp$seq_file #altering its format so row names are the sample ids
tax_tab<-tax_table(tax)
#phyloseq is essentially a useful bunch of functions for making plots and conducting analyses with this sort of data
ps <- phyloseq(otu, samp, tax_tab) #making the phyloseq object

library(tidyverse) ; packageVersion("tidyverse") # 1.3.1
library(phyloseq) ; packageVersion("phyloseq") # 1.22.3
library(vegan) ; packageVersion("vegan") # 2.5.4
library(viridis) ; packageVersion("viridis") # 0.5.1
library(decontam)

seqtab <- makeSequenceTable(merged)
seqtab.nochim <- removeBimeraDenovo(seqtab, verbose=T) # Identified 1331 bimeras out of 5895 input sequences.

asv_seqs <- colnames(seqtab.nochim)
asv_headers <- vector(dim(seqtab.nochim)[2], mode="character")

for (i in 1:dim(seqtab.nochim)[2]) {
  asv_headers[i] <- paste(">ASV", i, sep="_")
}

# making and writing out a fasta of our final ASV seqs:
asv_fasta <- c(rbind(asv_headers, asv_seqs))
write(asv_fasta, "ASVs.fa")

# count table:
asv_tab <- t(seqtab.nochim)
row.names(asv_tab) <- sub(">", "", asv_headers)
write.table(asv_tab, "ASVs_counts.tsv", sep="\t", quote=F, col.names=NA)

asv_tab<-table.no.chimeras
asv_headers<-paste("ASV_",1:ncol(asv_tab))
asv_seqs<-colnames(asv_tab)
colnames(asv_tab)<-asv_headers

library(DECIPHER)
download.file(url="http://www2.decipher.codes/Classification/TrainingSets/SILVA_SSU_r138_2019.RData", destfile="SILVA_SSU_r138_2019.RData")

## loading reference taxonomy object
load("SILVA_SSU_r138_2019.RData")

## loading DECIPHER
library(DECIPHER)
library(dada2)

## creating DNAStringSet object of our ASVs
dna <- DNAStringSet(getSequences(seqtab.nochim))

## and classifying
tax_info <- IdTaxa(test=dna, trainingSet=trainingSet, strand="both", processors=NULL)

# tax table:
# creating table of taxonomy and setting any that are unclassified as "NA"
ranks <- c("domain", "phylum", "class", "order", "family", "genus", "species")
asv_tax <- t(sapply(tax_info, function(x) {
  m <- match(ranks, x$rank)
  taxa <- x$taxon[m]
  taxa[startsWith(taxa, "unclassified_")] <- NA
  taxa
}))
colnames(asv_tax) <- ranks
rownames(asv_tax) <- gsub(pattern=">", replacement="", x=asv_headers)

write.table(asv_tax, "ASVs_taxonomy.tsv", sep = "\t", quote=F, col.names=NA)

colnames(asv_tab) 

contamdf.prev <- isContaminant(ps, method="prevalence", neg="negative")
table(contamdf.prev$contaminant)

#contam_df <- isContaminant(t(asv_tab), neg=metadata$negative) 

table(contamdf.prev$contaminant)
contam_asvs <- row.names(contamdf.prev[contamdf.prev$contaminant == TRUE, ])

tax[row.names(asv_tab) %in% contam_asvs, ]

asv_tab_no_contam <- asv_tab[!row.names(asv_tab) %in% contam_asvs, ]
asv_tax_no_contam <- tax[(!row.names(tax) %in% contam_asvs), ]


#asv_tab_no_contam<-asv_tab[asv_tab["primers-trimmed-GMCF-1240-Negative_S169_L001_R1_001.fastq",]==0,]
#asv_tax_no_contam<-asv_tax_no_contam[row.names(asv_tab_no_contam),]


asv_seq_headers_bind<-cbind(asv_seqs,row.names(asv_headers))
colnames(blorp)[7]<-"SEQ"
rownames(blorp)<-colnames(asv_tab)
asv_tax<-blorp


write(asv_fasta, "ASVs-no-contam.fa")
write.table(asv_tab_no_contam, "ASVs_counts-no-contam.tsv",
            sep="\t", quote=F, col.names=NA)
write.table(asv_tax_no_contam, "ASVs_taxonomy-no-contam.tsv",
            sep="\t", quote=F, col.names=NA)


write(summary_tab, "sample_info.tsv")

count_tab <- read.table("ASVs_counts-no-contam.tsv", header=T, row.names=1,
                        check.names=F, sep="\t")[ , -c(1:4)]

tax_tab <- as.matrix(read.table("ASVs_taxonomy-no-contam.tsv", header=T,
                                row.names=1, check.names=F, sep="\t"))


write.table(metadata, file = "sample_info.tsv", row.names=FALSE, sep="\t")


sample_info_tab <- read.table("sample_info.tsv", header=T, row.names=1,
                              check.names=F, sep="\t")                    

# and setting the color column to be of type "character", which helps later
sample_info_tab$color<- as.character(sample_info_tab$color)

sample_info_tab

#making sure species group is seen as a factor with levels
ps@sam_data$wgs.class <- factor(ps@sam_data$wgs.class, levels = c("BCCH", "HYCH", "ADV.CA", "CACH"))


#removing Chloroplast and mitochondria
ps_nochloro <- ps %>%
  subset_taxa(
    Kingdom == "Bacteria" &                   #only bacteria
      Family  != "mitochondria" &             #filter out mitochondria
      Class   != "Chloroplast"                #filter out chloroplasts
  )

```

subsetting time points and species groups 
```{r}

#seperating the aims 
initial_ps <- subset_samples(ps, sample_names(ps) %in% sample_names(ps)[1:90])
selffeeding_ps <- subset_samples(ps, sample_names(ps) %in% sample_names(ps)[91:147])
presurgery_ps <- subset_samples(ps, sample_names(ps) %in% sample_names(ps)[148:220])


#finding total reads 
initial_otu_table <- otu_table(initial_ps)
selffeed_otu_table <- otu_table(selffeeding_ps)
presurg_otu_table <- otu_table(presurgery_ps)

# Sum the counts across all samples
initial_total_reads <- sum(initial_otu_table)
selffeed_total_reads <- sum(selffeed_otu_table)
presurg_total_reads <- sum(presurg_otu_table)

# Print the total number of reads
print(initial_total_reads)
print(selffeed_total_reads)
print(presurg_total_reads)


# finding average number of reads per bird
otu_table <- otu_table(presurgery_ps)
total_reads <- sum(otu_table)
num_samples <- ncol(otu_table)
average_reads_per_sample <- total_reads / num_samples
print(average_reads_per_sample)



#ALL The same below ...
#finding total number of phyla and genus for each species group at each collection time 
bcch_tax_table <- tax_table(BCCH_ps)
selffeed_tax_table <- tax_table(selffeeding_ps)
presurg_tax_table <- tax_table(presurgery_ps)

# Extract the phylum column
initial_phyla <- bcch_tax_table[, "Phylum"]
selffeed_phyla <- selffeed_tax_table[, "Genus"]
presurg_phyla <- presurg_tax_table[, "Genus"]

# Count the unique phylum names
total_initial_phyla <- length(unique(initial_phyla))
total_selffeed_phyla <- length(unique(selffeed_phyla))
total_presurg_phyla <- length(unique(presurg_phyla))

# Print the total number of phyla
print(total_initial_phyla)
print(total_selffeed_phyla)
print(total_presurg_phyla)

```


average all the individuals in each group and in each nest_site to see initial species and geographical GM variation 
*insert different collection time points for analysis 
```{r}
library(phyloseq)
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# Define desired orders
desired_order_wgs <- c("BCCH", "HYCH", "ADV.CA", "CACH") # Replace with your actual groups
desired_order_nest <- c("HickoryRun", "TwinLakes", "Tuscarora", "Jacobsburg", "SouthMtn", "Desales", "Nockamixon", "PeaceValley", "Nolde") 

# Add Deinococcota to the fixed color scheme
fixed_colors <- c(brewer.pal(12, "Set3"), brewer.pal(8, "Dark2"), "gray")
names(fixed_colors) <- c("Firmicutes", "Proteobacteria", "Bacteroidota", "Bdellovibrionota", "Actinobacteriota", 
                         "Fusobacteriota", "Cyanobacteria", "Chloroflexi", "Verrucomicrobiota", "Planctomycetota", 
                         "Gemmatimonadota", "Patescibacteria", "Dependentiae","Deinococcota", "Nitrospirota", "Myxococcota", 
                         "Acidobacteriota", "Desulfobacterota", "Other")


# Use psmelt to convert phyloseq object to a data frame
melted_df <- psmelt(selffeeding_ps_nochloro)

# Filter out samples with NA in wgs.class or Nest_Site
melted_df <- melted_df %>%
  filter(!is.na(wgs.class) & !is.na(Nest_Site))
#factor based on the levels
melted_df$wgs.class <- factor(melted_df$wgs.class, levels = desired_order_wgs)
melted_df$Nest_Site <- factor(melted_df$Nest_Site, levels = desired_order_nest)

# Calculate the mean relative abundance for each phylum
mean_abund <- melted_df %>%
  group_by(Phylum) %>%
  summarize(MeanAbundance = mean(Abundance, na.rm = TRUE)) %>%
  arrange(desc(MeanAbundance))

# Identify the top 10 most abundant phyla
top10_phyla <- mean_abund$Phylum[1:10]

# Modify the data frame to group all other phyla into "Other"
melted_df <- melted_df %>%
  mutate(Phylum = ifelse(Phylum %in% top10_phyla, Phylum, "Other"))

# Summarize the data by wgs.class and phylum
abund_long_wgs <- melted_df %>%
  group_by(wgs.class, Phylum) %>%
  summarize(Abundance = mean(Abundance)) %>%
  ungroup()

# Create the bar plot for wgs.class
topphyla_barplot_wgs <- ggplot(abund_long_wgs, aes(x = wgs.class, y = Abundance, fill = Phylum)) +
  geom_col(position = "fill") +
  labs(x = "wgs.class", y = "% of 16S rRNA gene copies recovered", 
       title = "Relative Abundance of Top 10 Phyla and Others by wgs.class") +
  scale_fill_manual(values = fixed_colors, name = "Phylum", breaks = c(top10_phyla, "Other")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

print(topphyla_barplot_wgs)

# Save the plot to a PDF
pdf("/Users/russ/Desktop/topphyla_barplot_wgs.pdf", width = 15, height = 10)
print(topphyla_barplot_wgs)
dev.off()

# Summarize the data by Nest_Site and phylum
abund_long_nest <- melted_df %>%
  group_by(Nest_Site, Phylum) %>%
  summarize(Abundance = mean(Abundance)) %>%
  ungroup()

# Create the bar plot for Nest_Site
topphyla_barplot_nest <- ggplot(abund_long_nest, aes(x = Nest_Site, y = Abundance, fill = Phylum)) +
  geom_col(position = "fill") +
  labs(x = "Nest_Site", y = "% of 16S rRNA gene copies recovered", 
       title = "Relative Abundance of Top 10 Phyla and Others by Nest_Site") +
  scale_fill_manual(values = fixed_colors, name = "Phylum", breaks = c(top10_phyla, "Other")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

print(topphyla_barplot_nest)

# Save the plot to a PDF
pdf("/Users/russ/Desktop/topphyla_barplot_nest.pdf", width = 15, height = 10)
print(topphyla_barplot_nest)
dev.off()

```

alpha and beta diversity testing 
```{r}
library(vegan)
library(dplyr)

#alpha diversity measures
shannon_diversity<-estimate_richness(ps_nochloro, measures = "shannon")
chao1<- estimate_richness(ps_nochloro, measures = "Chao1")
richness<- estimate_richness(ps_nochloro, measures = "Observed")

# Bray-Curtis dissimilarity
bray <- phyloseq::distance(ps_nochloro, method = "bray")

# Jaccard dissimilarity
jaccard <- phyloseq::distance(ps_nochloro, method = "jaccard")

#perform PERMANOVA
permanova_result <- adonis2(bray ~ wgs.class+collection_time+Collection_year+Sex, data = meta_data, permutations = 10000, na.action = "na.exclude", by = "margin")
print(permanova_result)

#perform PERMANOVA
permanova_result <- adonis2(jaccard ~ wgs.class+collection_time+Collection_year+Sex, data = meta_data, permutations = 10000, na.action = "na.exclude", by = "margin", strata = meta_data$band_id)
print(permanova_result)


```

graph for alpha diversity differences
*input diff aspects of alpha diversity for graphing 
```{r}
#cleaner graph for collection time on x axis and alpha diversity on the y axis 
library(ggplot2)
library(cowplot)

asvnum1 <- ggplot(diversity_df, aes(x = new_wgs, y = Shannon, fill = new_wgs)) +
  geom_boxplot() +
  labs(
    x = "Ancestry",
    y = "Chao Diversity",
    fill = "Ancestry",
    title = "Distribution of ASVs by Ancestry"
  ) +
  theme_minimal(base_size = 14) + # Clean, modern theme with a slightly larger font size
  theme(
    plot.title = element_text(
      hjust = 0.5, face = "bold", size = 16, color = "black", margin = margin(b = 10)
    ), # Title styling
    plot.title.position = "plot", # Ensures the title is placed at the top
    plot.title.background = element_rect(
      fill = "lightblue", color = "black", linewidth = 1
    ), # Adds a box behind the title
    panel.border = element_rect(color = "black", fill = NA, linewidth = 1), # Outline around the graph
    axis.title = element_text(face = "bold"),
    axis.text = element_text(color = "black"),
    legend.position = "top", # Moves legend to the top for better space management
    legend.background = element_rect(color = "black", fill = "white", linewidth = 0.5),
    legend.box.margin = margin(t = 5, b = 5)
  ) +
  scale_fill_brewer(palette = "Set2")+
  theme_cowplot(12)

print(asvnum1)

pdf(paste("/Users/russ/Desktop/asvnum1.pdf"), width = 10, height = 10)
print(asvnum1)
dev.off()
```


PCoA graphing 
*input diff distance matrix for each graph
```{r}
# Load necessary libraries
library(phyloseq)
library(ggplot2)
library(vegan)  # For PCoA analysis
library(ggpubr) # For adding ellipses

# Extract the sample data
sample_data_df <- as.data.frame(sample_data(ps_nochloro))

# Remove rows with NAs in collection_time or wgs.class
sample_data_df <- sample_data_df[!is.na(sample_data_df$collection_time) & !is.na(sample_data_df$wgs.class), ]

# Subset the phyloseq object to remove samples with NA values
ps_nochloro_filtered <- prune_samples(rownames(sample_data_df), ps_nochloro)

# Perform PCoA analysis on the filtered data
ordination <- ordinate(ps_nochloro_filtered, method = "PCoA", distance = "bray")

point_colors <- c("initial" = "", "self_feeding" = "", "pre_surgery" = "")
fill_colors <- c("BCCH" = "aquamarine3", "HYCH" = "chocolate1", "ADV.CA" = "cornflowerblue", "CACH" = "pink3")

# Create the PCoA plot
pcoa_plot <- plot_ordination(ps_nochloro, ordination, type = "samples") +
  geom_point(aes(shape = collection_time, color = new_wgs), size = 3) +  
  stat_ellipse(aes(group = new_wgs, fill = new_wgs), 
               geom = "polygon", alpha = 0.2) +   
  #scale_fill_manual(values = fill_colors) +  # Provide colors for ellipses
  #scale_color_manual(values = fill_colors) + # Provide colors for points
  theme_minimal() +
  labs(title = "PCoA of Microbial Communities",
       shape = "Collection Time", color = "Collection Time")

# Display the plot
print(pcoa_plot)


# Save the plot to a PDF
pdf("/Users/russ/Desktop/pcoa_plot.pdf", width = 15, height = 10)
print(pcoa_plot)
dev.off()

```

chloroplast subsetting 
```{r}
library(phyloseq)
library(dplyr)

  # Convert tax_table to a data frame
  tax_df <- as.data.frame(tax_table(ps))
  
  # Identify ASVs that belong to the "Chloroplast" order
  chloroplast_asvs <- rownames(tax_df[tax_df$Order == "Chloroplast", ])
  
  # Subset the phyloseq object using these ASVs
  chloro_ps <- prune_taxa(chloroplast_asvs, ps)


# Inspect the subsetted object
print(chloro_ps)


#subsetting the new chloroplast phyloseq to just the specific time points
initial_chloro_ps <- subset_samples(chloro_ps, chloro_ps@sam_data$collection_time == "initial")
selffeed_chloro_ps <- subset_samples(chloro_ps, chloro_ps@sam_data$collection_time == "self_feeding")
presurg_chloro_ps <- subset_samples(chloro_ps, chloro_ps@sam_data$collection_time == "pre_surgery")


```


grpahing chloroplast data
```{r}
library(phyloseq)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RColorBrewer)

# Convert phyloseq object to data frame
abund_df <- as.data.frame(otu_table(initial_chloro_ps))
meta_data <- as(sample_data(initial_chloro_ps), "data.frame")

# Combine abundance data with metadata
abund_df <- cbind(SampleID = rownames(abund_df), abund_df)

#fix the seq_file name to also be SampleID
meta_data <- as(sample_data(initial_chloro_ps), "data.frame")
colnames(meta_data)[colnames(meta_data) == "seq_file"] <- "SampleID"

# Convert abundance data to long format and join with metadata
abund_long <- abund_df %>%
  pivot_longer(-SampleID, names_to = "ASV", values_to = "Abundance") %>%
  merge(meta_data, by = "SampleID")

# Calculate relative abundance
abund_long <- abund_long %>%
  group_by(SampleID) %>%
  mutate(RelativeAbundance = Abundance / sum(Abundance, na.rm = TRUE)) %>%
  ungroup()

# Summarize relative abundance by wgs.class and Nest_Site
abund_wgs <- abund_long %>%
  group_by(new_wgs, ASV) %>%
  summarize(RelativeAbundance = mean(RelativeAbundance, na.rm = TRUE)) %>%
  arrange(new_wgs, desc(RelativeAbundance))

abund_nest <- abund_long %>%
  group_by(Nest_Site, ASV) %>%
  summarize(RelativeAbundance = mean(RelativeAbundance, na.rm = TRUE)) %>%
  arrange(Nest_Site, desc(RelativeAbundance))

# Define the number of top ASVs to select
top_n <- 10

# Top ASVs for each wgs.class
top_asvs_wgs <- abund_wgs %>%
  group_by(new_wgs) %>%
  slice_max(order_by = RelativeAbundance, n = top_n, with_ties = FALSE) %>%
  ungroup()

# Top ASVs for each Nest_Site
top_asvs_nest <- abund_nest %>%
  group_by(Nest_Site) %>%
  slice_max(order_by = RelativeAbundance, n = top_n, with_ties = FALSE) %>%
  ungroup()

# Create a lookup table for ASVs to shorter names
asv_lookup_nest <- data.frame(
  ASV = unique(top_asvs_nest$ASV),
  ShortName = paste0("ASV", seq_along(unique(top_asvs_nest$ASV)))
)

# Create a lookup table for ASVs to shorter names
asv_lookup_wgs <- data.frame(
  ASV = unique(top_asvs_wgs$ASV),
  ShortName = paste0("ASV", seq_along(unique(top_asvs_wgs$ASV)))
)

# Join the lookup table with the top ASVs data
top_asvs_nest <- top_asvs_nest %>%
  left_join(asv_lookup_nest, by = "ASV")

# Join the lookup table with the top ASVs data
top_asvs_wgs <- top_asvs_wgs %>%
  left_join(asv_lookup_wgs, by = "ASV")


# Plot top ASVs for each Nest_Site based on relative abundance
wgs.class_chloro_asv<-ggplot(top_asvs_wgs, aes(x = reorder(ShortName, RelativeAbundance), y = RelativeAbundance, fill = new_wgs)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ new_wgs, scales = "free_x") +
  labs(x = "ASV", y = "Relative Abundance", title = "Top 10 ASVs for Each Species Group (Relative Abundance)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_fill_brewer(palette = "Set3")

print(wgs.class_chloro_asv)

# Save the plot to a PDF
pdf("/Users/russ/Desktop/wgs.class_chloro_asv.pdf", width = 15, height = 30)
print(wgs.class_chloro_asv)
dev.off()

# Plot top ASVs for each Nest_Site based on relative abundance
nest_chloro_asv<-ggplot(top_asvs_nest, aes(x = reorder(ShortName, RelativeAbundance), y = RelativeAbundance, fill = Nest_Site)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Nest_Site, scales = "free_x") +
  labs(x = "ASV", y = "Relative Abundance", title = "Top 10 ASVs for Each Nest_Site (Relative Abundance)") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  scale_fill_brewer(palette = "Set3")
print(nest_chloro_asv)

# Save the plot to a PDF
pdf("/Users/russ/Desktop/nest_chloro_asv.pdf", width = 15, height = 30)
print(nest_chloro_asv)
dev.off()

```



running PERMANOVA stats chloro phyloseq 
```{r}
#PERMANOVA of beta diversity ordination plot 
library(vegan)

#set meta_data
meta_data_chloro<-sample_data(chloro_ps)
meta_data_chloro <- data.frame(as(meta_data, "matrix"))


# Bray-Curtis dissimilarity
bray_chloro <- phyloseq::distance(chloro_ps, method = "bray")

# Jaccard dissimilarity
jaccard_chloro <- phyloseq::distance(chloro_ps, method = "jaccard")

# Add a small value to all zero entries
bray_chloro[bray_chloro == 0] <- 0.0001
bray_chloro[is.na(bray_chloro)] <- 0.0001

jaccard_chloro[jaccard_chloro == 0] <- 0.0001
jaccard_chloro[is.na(jaccard_chloro)] <- 0.0001

#perform PERMANOVA
permanova_result <- adonis2(jaccard_chloro ~ wgs.class+Sex+collection_time, data = meta_data_chloro, permutations = 10000, na.action = "na.exclude", by = "margin")
print(permanova_result)

```


alpha diversity of chloroplast 
```{r}
shannon_diversity<-estimate_richness(chloro_ps, measures = "shannon")
inverse_simpson_index <- estimate_richness(chloro_ps, measures = "InvSimpson")
chao1<- estimate_richness(chloro_ps, measures = "Chao1")
richness<- estimate_richness(chloro_ps, measures = "Observed")

# Calculate diversity measures
diversity_df_chloro <- data.frame(
  SampleID = sample_names(chloro_ps),
  Shannon = estimate_richness(chloro_ps, measures = "Shannon")[, "Shannon"],
  Chao = estimate_richness(chloro_ps, measures = "Chao1")[, "Chao1"],
  Richness = estimate_richness(chloro_ps, measures = "Observed")[, "Observed"],
  OTUs = otu_table(chloro_ps) %>% rowSums()
)

#making the metadata object and confirming dataframes 
meta_data_chloro<-sample_data(chloro_ps)
meta_data_chloro <- data.frame(as(meta_data, "matrix"))
diversity_df_chloro<-as.data.frame(diversity_df_chloro)

# Rename seq_file to SampleID in meta_data
colnames(meta_data_chloro)[colnames(meta_data_chloro) == "seq_file"] <- "SampleID"

# Combine with metadata
diversity_df_chloro <- merge(diversity_df_chloro, meta_data_chloro, by = "SampleID")

# Filter out NAs
diversity_df_chloro <- diversity_df_chloro %>%
  filter(!is.na(wgs.class) & !is.na(collection_time))

# Set desired order for wgs.class and collection_time
desired_order_wgs <- c("BCCH", "HYCH", "ADV.CA", "CACH") 
diversity_df_chloro$wgs.class <- factor(diversity_df_chloro$wgs.class, levels = desired_order_wgs)
desired_order_collection_time <- c("initial", "self_feeding", "pre_surgery") 
diversity_df_chloro$collection_time <- factor(diversity_df_chloro$collection_time, levels = desired_order_collection_time)



#making sure the variables are seen as factors not characters 
diversity_df_chloro$wgs.class<-as.factor(diversity_df_chloro$wgs.class)
diversity_df_chloro$collection_time<-as.factor(diversity_df_chloro$collection_time)
diversity_df_chloro$Collection_year<-as.factor(diversity_df_chloro$Collection_year)
diversity_df_chloro$Sex<-as.factor(diversity_df_chloro$Sex)
diversity_df_chloro$Nest_Site<-as.factor(diversity_df_chloro$Nest_Site)
diversity_df_chloro$Nest<-as.factor(diversity_df_chloro$Nest)




#SHANNON DIVERSITY
#simplified version
shan.init <- lmer(Shannon ~ wgs.class*Sex*collection_time + (1|Collection_year/Nest), data = diversity_df_chloro)
summary(shan.init)
#check the Anova 
Anova(shan.init)
shan.init.ms <- update(shan.init, ~ . 
  - wgs.class:Sex
  -wgs.class:collection_time
  -Sex:collection_time
  -wgs.class:Sex:collection_time)
anova(shan.init.ms,shan.init)
Anova(shan.init.ms)

#CHAO DIVERSITY
#simplified version
chao.init <- lmer(Chao ~ wgs.class*Sex*collection_time + (1|Collection_year/Nest), data = diversity_df_chloro)
summary(chao.init)
#check the Anova 
Anova(chao.init)
chao.init.ms <- update(chao.init, ~ . 
  - wgs.class:Sex
  -wgs.class:collection_time
  -Sex:collection_time
  -wgs.class:Sex:collection_time)
anova(chao.init.ms,chao.init)
Anova(chao.init.ms)

#OBSERVED RICHNESS
#simplified version
rich.full <- lmer(Richness ~ wgs.class*Sex*collection_time + (1|Collection_year/Nest), data = diversity_df_chloro)
summary(rich.full)
#check the Anova 
Anova(rich.full)


#NUMBER OF ASVs
#simplified version
asvnum.init <- lmer(OTUs ~ wgs.class*Sex*collection_time + (1|Collection_year/Nest), data = diversity_df_chloro)
summary(asvnum.init)
#check the Anova 
Anova(asvnum.init)
asvnum.init.ms <- update(asvnum.init, ~ . 
  - wgs.class:Sex
  -wgs.class:collection_time
  -Sex:collection_time
  -wgs.class:Sex:collection_time)
anova(asvnum.init.ms,asvnum.init)
Anova(asvnum.init.ms)




```


post hoc tests for the chloroplast data ALPHA
```{r}
library(emmeans)
# checking post hoc after the linear mixed effects above 

# Visualize interaction while accounting for nesting
#shan_emm<-emmip(shan.init, ~ wgs.class | Nest_Site)

# Get the estimated marginal means for the main effects and interactions
emmeans_results <- emmeans(chao.init, pairwise ~ wgs.class)

# Perform pairwise comparisons (adjust method can be changed as needed)
contrast_results <- pairs(emmeans_results, adjust = "tukey")

# Print the results
summary(contrast_results)

```

post hoc tests for the chloroplast data BETA
```{r}
library(vegan)
#set meta_data
meta_data<-sample_data(presurg_chloro_ps)
meta_data <- data.frame(as(meta_data, "matrix"))


# Bray-Curtis dissimilarity
bray <- phyloseq::distance(chloro_ps, method = "bray")

# Jaccard dissimilarity
jaccard <- phyloseq::distance(chloro_ps, method = "jaccard")

# Add a small value to all zero entries
bray[bray == 0] <- 0.0001
bray[is.na(bray)] <- 0.0001

jaccard[jaccard == 0] <- 0.0001
jaccard[is.na(jaccard)] <- 0.0001



meta_data_clean <- meta_data[!is.na(meta_data$collection_time), ]

jaccard_clean <- as.dist(as.matrix(jaccard)[rownames(meta_data_clean), rownames(meta_data_clean)])

bray_clean <- as.dist(as.matrix(bray)[rownames(meta_data_clean), rownames(meta_data_clean)])

# Assuming jaccard_clean is the distance matrix and collection_time is the grouping variable
dispersion_test <- betadisper(jaccard_clean, meta_data_clean$Collection_year)

dispersion_result <- anova(dispersion_test)
print(dispersion_result)

pairwise_dispersion <- TukeyHSD(dispersion_test)
print(pairwise_dispersion)

```


differential abundance of top 10 ASVs using ANCOMBC 
*insert each different subset of groupings to find significant ASVs for each ancestry group or collection time 
```{r}
library(ANCOMBC)
library(phyloseq)


# Subset top 10 ASVs
top10_asvs <- head(order(-colSums(abund_df)), 10)
top10_ps <- subset_taxa(initial_ps, taxa_names(initial_ps) %in% top10_asvs)

library(ANCOMBC)


# Run ANCOMBC analysis
full_ancombc_collectime <- ancombc2(data = ps_nochloro, tax_level = "Genus", fix_formula = "collection_time", p_adj_method = "holm", group = "collection_time", global = TRUE)

print(full_ancombc_collectime$res_global)

# Extract significant ASVs based on q-value threshold
significant_asvs <- rownames(full_ancombc_collectime$res_global)[full_ancombc_collectime$res_global$q_val < 0.05]
print(significant_asvs)

# Extract q-values for the significant ASVs using row names
significant_q_vals <- full_ancombc_collectime$res_global$q_val[rownames(full_ancombc_collectime$res_global) %in% significant_asvs]

# Extract the rows corresponding to significant ASVs
res_global_stats <- full_ancombc_collectime$res_global[rownames(full_ancombc_collectime$res_global) %in% significant_asvs, ]

# Export the results to a CSV file
write.csv(res_global_stats, "/Users/russ/Desktop/significant_asvs_with_taxonomy.csv", row.names = FALSE)

# Print the data frame to check the output
print(res_global_stats)

```


finding the significant asv and putting them in their own object for a differential abundance graph 
```{r}

ancomb_ps<-ps_nochloro

#access and modify the sample data
ancomb_ps@sam_data$wgs.class <- as.factor(ancomb_ps@sam_data$wgs.class)

# filter out samples with NA in wgs.class
filtered_samples_wgs <- !is.na(sample_data(ancomb_ps)$wgs.class)

# prune the phyloseq object to only keep samples without NA in wgs.class
ancomb_ps_filtered_wgs <- prune_samples(filtered_samples_wgs, ancomb_ps)

# run ANCOMBC analysis across the wgs.class groups
full_ancombc_wgs <- ancombc2(
  data = ancomb_ps_filtered_wgs, 
  tax_level = "Genus", 
  fix_formula = "wgs.class",  # Change to compare wgs.class groups
  rand_formula = "(1 | collection_time)",  # You can keep collection_time in the random formula if it's relevant
  p_adj_method = "holm", 
  pseudo = 0, 
  pseudo_sens = TRUE, 
  group = "wgs.class",  # Change to group by wgs.class
  global = TRUE, 
  pairwise = TRUE, 
  dunnet = TRUE
)



library(ggplot2)
library(dplyr)
library(tidyr)

# Set to dataframe
df_results <- as.data.frame(full_ancombc_wgs$res)  # Use the correct ANCOMBC result

# Reshape data for ggplot
df_results_long <- df_results %>%
  pivot_longer(
    cols = starts_with("lfc_"),  # Selecting log fold change columns
    names_to = "comparison",  # Creating a new column for comparison (based on wgs.class)
    names_prefix = "lfc_",  # Removing the "lfc_" prefix from names
    values_to = "log_fold_change"  # The values are the log fold changes
  ) %>%
  pivot_longer(
    cols = starts_with("se_"),
    names_to = "se_comparison",
    names_prefix = "se_",  # Adjust for the standard error prefix
    values_to = "se"
  ) %>%
  pivot_longer(
    cols = starts_with("p_"),
    names_to = "p_comparison",
    names_prefix = "p_",  # Adjust for the p-value prefix
    values_to = "p_value"
  ) %>%
  filter(comparison == se_comparison & comparison == p_comparison) %>%  # Matching comparisons across columns
  mutate(
    significance = case_when(
      p_value < 0.001 ~ "***",
      p_value < 0.01 ~ "**",
      p_value < 0.05 ~ "*",
      TRUE ~ ""
    ),
    lower_ci = log_fold_change - 1.96 * se,  # Calculating 95% CI
    upper_ci = log_fold_change + 1.96 * se
  )



library(dplyr)
library(tidyr)
library(ggplot2)

# Select relevant columns from df_results for wgs.class comparisons
df_comparison <- df_results %>%
    dplyr::select(taxon, 
                  lfc_wgs.classHYCH, 
                  lfc_wgs.classBCCH,
                  lfc_wgs.classCACH,
                  se_wgs.classHYCH, 
                  se_wgs.classBCCH, 
                  se_wgs.classCACH, 
                  passed_ss_wgs.classHYCH, 
                  passed_ss_wgs.classBCCH, 
                  passed_ss_wgs.classCACH)

# Create df_fig_comparison by pivoting the necessary columns
df_fig_comparison <- df_comparison %>%
    pivot_longer(
        cols = starts_with("lfc_wgs.class"),  # Pivot the log fold changes for wgs.class
        names_to = "comparison", 
        names_prefix = "lfc_wgs.class",  # Remove prefix to clean up comparison names
        values_to = "log_fold_change"
    ) %>%
    pivot_longer(
        cols = starts_with("se_wgs.class"),  # Pivot the standard errors for wgs.class
        names_to = "se_comparison", 
        names_prefix = "se_wgs.class",  # Remove prefix to clean up SE comparison names
        values_to = "se"
    ) %>%
    # Filter to match the comparison in log fold change and SE
    filter(comparison == se_comparison) %>%
    # Add passed significance status for each comparison
    mutate(
        passed_ss = case_when(
            comparison == "HYCH" ~ passed_ss_wgs.classHYCH,
            comparison == "BCCH" ~ passed_ss_wgs.classBCCH,
            comparison == "CACH" ~ passed_ss_wgs.classCACH,
            TRUE ~ NA
        ),
        direct = ifelse(log_fold_change > 0, "Positive LFC", "Negative LFC"),
        color = ifelse(passed_ss == TRUE, "aquamarine3", "black")
    ) %>%
    arrange(desc(log_fold_change))

# Create the plot with different wgs.class comparisons
fig_comparison <- ggplot(df_fig_comparison, aes(x = reorder(taxon, log_fold_change), y = log_fold_change, fill = direct)) + 
    geom_bar(stat = "identity", width = 0.7, color = "black", 
             position = position_dodge(width = 0.4)) +
    geom_errorbar(aes(ymin = log_fold_change - se, ymax = log_fold_change + se), 
                  width = 0.2, position = position_dodge(0.4), color = "black") + 
    facet_wrap(~ comparison, scales = "free", ncol = 1) +  # Separate panels for each wgs.class comparison
    labs(x = NULL, y = "Log Fold Change", 
         title = "Log Fold Changes for WGS Class Comparisons (HYCH, ADV.CA, CACH)") + 
    scale_fill_discrete(name = NULL) +
    theme_bw() + 
    theme(
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1, color = df_fig_comparison$color)
    ) +
    coord_cartesian(ylim = c(-4, 4))  # Set y-axis limits

# Display the figure
print(fig_comparison)

# Save the plot to a PDF
pdf("/Users/russ/Desktop/fig_comparison_wgs_class.pdf", width = 10, height = 20)
print(fig_comparison)
dev.off()


```



differential abundance graphs 
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Select relevant columns from df_results for wgs.class comparisons, including p-values
df_comparison <- df_results %>%
    dplyr::select(taxon, 
                  lfc_wgs.classHYCH, 
                  lfc_wgs.classBCCH,
                  lfc_wgs.classCACH,
                  se_wgs.classHYCH, 
                  se_wgs.classBCCH, 
                  se_wgs.classCACH, 
                  p_wgs.classHYCH,
                  p_wgs.classBCCH,
                  p_wgs.classCACH)

# Create df_fig_comparison by pivoting the necessary columns
df_fig_comparison <- df_comparison %>%
    pivot_longer(
        cols = starts_with("lfc_wgs.class"),  # Pivot the log fold changes for wgs.class
        names_to = "comparison", 
        names_prefix = "lfc_wgs.class",  # Remove prefix to clean up comparison names
        values_to = "log_fold_change"
    ) %>%
    pivot_longer(
        cols = starts_with("se_wgs.class"),  # Pivot the standard errors for wgs.class
        names_to = "se_comparison", 
        names_prefix = "se_wgs.class",  # Remove prefix to clean up SE comparison names
        values_to = "se"
    ) %>%
    pivot_longer(
        cols = starts_with("p_wgs.class"),  # Pivot the p-values for wgs.class
        names_to = "p_comparison", 
        names_prefix = "p_wgs.class",  # Remove prefix to clean up p-value names
        values_to = "p_value"
    ) %>%
    # Filter to match the comparison in log fold change, SE, and p-value
    filter(comparison == se_comparison & comparison == p_comparison) %>%
    # Add passed significance status for each comparison
    mutate(
        passed_ss = case_when(
            comparison == "HYCH" ~ p_value < 0.05,
            comparison == "BCCH" ~ p_value < 0.05,
            comparison == "CACH" ~ p_value < 0.05,
            TRUE ~ NA
        ),
        direct = ifelse(log_fold_change > 0, "Positive LFC", "Negative LFC"),
        color = ifelse(passed_ss == TRUE, "aquamarine3", "black")
    ) %>%
    arrange(desc(log_fold_change))

# Filter for significant p-values (p < 0.05)
df_significant <- df_fig_comparison %>%
    filter(passed_ss == TRUE)

# View the significant genera
significant_genera <- df_significant %>%
    dplyr::select(taxon, comparison, log_fold_change, p_value) %>%
    arrange(comparison, log_fold_change)

# Print the significant genera
print(significant_genera)

# Optional: Create a plot highlighting the significant genera
fig_significant <- ggplot(df_significant, aes(x = reorder(taxon, log_fold_change), y = log_fold_change, fill = direct)) + 
    geom_bar(stat = "identity", width = 0.7, color = "black", 
             position = position_dodge(width = 0.4)) +
    geom_errorbar(aes(ymin = log_fold_change - se, ymax = log_fold_change + se), 
                  width = 0.2, position = position_dodge(0.4), color = "black") + 
    facet_wrap(~ comparison, scales = "free", ncol = 1) +  # Separate panels for each wgs.class comparison
    labs(x = NULL, y = "Log Fold Change", 
         title = "Significant Genera in WGS Class Comparisons (HYCH, BCCH, CACH)") + 
    scale_fill_discrete(name = NULL) +
    theme_bw() + 
    theme(
        plot.title = element_text(hjust = 0.5),
        panel.grid.minor.y = element_blank(),
        axis.text.x = element_text(angle = 60, hjust = 1, color = df_significant$color)
    ) +
    coord_cartesian(ylim = c(-4, 4))  # Set y-axis limits

# Display the plot
print(fig_significant)

# Optional: Save the significant plot to a PDF
pdf("/Users/russ/Desktop/significant_genera_wgs_class.pdf", width = 10, height = 20)
print(fig_significant)
dev.off()

```


making a graph with the ancombc results 
```{r}
# Load necessary libraries
library(dplyr)

# Extract OTU and taxonomy tables
abundance_matrix <- otu_table(presurgery_ps_nochloro)
taxonomy_df <- tax_table(presurgery_ps_nochloro)

# Convert OTU table to data frame and rename columns to ASV_1, ASV_2, etc.
abundance_df <- as.data.frame(abundance_matrix)
asv_seqs <- colnames(abundance_df)

# Create ASV headers like ASV_1, ASV_2
asv_headers <- paste0("ASV_", seq_along(asv_seqs))
colnames(abundance_df) <- asv_headers

# Convert taxonomy table to a data frame and move rownames (ASV names) to the first column
taxonomy_df <- as.data.frame(taxonomy_df)
asv_headers <- paste0("ASV_", seq_along(asv_seqs))
rownames(taxonomy_df) <- asv_headers

abundance_df<-t(abundance_df)

# Merge OTU and taxonomy tables (rows from taxonomy_df and columns from abundance_df)
abundance_taxa_df <- cbind(taxonomy_df, abundance_df)

# Ensure that abundance_taxa_df is a data frame
abundance_taxa_df <- as.data.frame(abundance_taxa_df)


# Extract the columns with ASV counts (assuming these are all columns except the taxonomy columns)
asv_columns <- colnames(abundance_taxa_df)[grepl("^primers-trimmed", colnames(abundance_taxa_df))]

# Convert these columns to numeric
abundance_taxa_df[asv_columns] <- lapply(abundance_taxa_df[asv_columns], as.numeric)

# Pivot longer to get sample-wise counts
abundance_long <- abundance_taxa_df %>%
  pivot_longer(cols = starts_with("primers-trimmed"), names_to = "Sample", values_to = "Abundance")

library(dplyr)

# Group by Genus and summarize total abundance across all samples
abundance_grouped_genus <- abundance_long %>%
  group_by(Genus) %>%
  summarize(Total_Abundance = sum(Abundance, na.rm = TRUE), .groups = 'drop')

# Group by Genus and Sample, then summarize the abundance
abundance_grouped_genus_sample <- abundance_long %>%
  group_by(Genus, Sample) %>%
  summarize(Abundance_Per_Sample = sum(Abundance, na.rm = TRUE), .groups = 'drop')

# View the result
head(abundance_grouped_genus)


# Identify the top 10 genera by total abundance, excluding NA genera
top_10_genera <- abundance_long %>%
  filter(!is.na(Genus)) %>%          # Exclude NA values in Genus
  group_by(Genus) %>%
  summarize(Total_Abundance = sum(Abundance)) %>%
  top_n(10, Total_Abundance) %>%
  pull(Genus)

# 1. Filter the data to include only the top 20 genera
abundance_top10 <- abundance_grouped_genus_sample %>%
  filter(Genus %in% top_10_genera)

colnames(abundance_top10)[colnames(abundance_top10) == "Abundance_Per_Sample"] <- "Abundance"


# 2. Calculate total abundance for each sample
total_abundance <- abundance_top10 %>%
  group_by(Sample) %>%
  summarize(Total_Abundance = sum(Abundance, na.rm = TRUE))

# 3. Calculate relative abundance
abundance_relative <- abundance_top10 %>%
  left_join(total_abundance, by = "Sample") %>%
  mutate(Relative_Abundance = Abundance / Total_Abundance)

# 4. Update metadata column name
metadata_df <- ps_nochloro@sam_data
colnames(metadata_df)[colnames(metadata_df) == "seq_file"] <- "Sample"

# Merge with metadata
full_data <- abundance_relative %>%
  left_join(metadata_df, by = "Sample")

# Optionally, pivot back to wide format for visualization or further analysis
abundance_relative_wide <- abundance_relative %>%
  pivot_wider(
    names_from = Sample,
    values_from = Relative_Abundance
  )

library(ggplot2)

full_data <- full_data %>% filter(!is.na(wgs.class))

genus_palette <- c(
  "Catellicoccus" = "#E41A1C", "Clostridium sensu stricto 1" = "#377EB8", "Corynebacterium" = "#4DAF4A",
  "Cronobacter" = "#984EA3", "Enterobacillus" = "#FF7F00", "Enterobacter" = "#FFED6F",
  "Enterococcus" = "#A65628", "Exiguobacterium" = "#F781BF", "Lactococcus" = "#CCEBC5",
  "Latilactobacillus" = "#66C2A5", "Mycoplasma" = "#FB9A99", "Pseudomonas" = "#FDB462", "Rickettsiella" = "#B3DE69", "Romboutsia" = "#FCCDE5", "Staphylococcus" = "#D9D9D9", "Ureaplasma" = "#CAB2D6", "Serratia" = "cyan2")


# Calculate total abundance globally (across all samples) for each Genus
genus_abundance_order <- full_data %>%
  group_by(Genus) %>%
  summarize(Total_Abundance = sum(Relative_Abundance, na.rm = TRUE)) %>%
  arrange(-Total_Abundance) %>%
  pull(Genus)

# Reorder Genus globally based on the total abundance across all samples
full_data <- full_data %>%
  mutate(Genus = factor(Genus, levels = genus_abundance_order))

# Create the stacked bar plot with globally ordered genera
top10genera <- ggplot(full_data, aes(x = Genus, y = Relative_Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  scale_fill_manual(values = genus_palette) +  # Apply the custom palette
  facet_wrap(~wgs.class) +
  labs(x = NULL, y = "% of 16S rRNA gene copies recovered", 
       title = "Top 20 Most Abundant Genera Across Samples") +
  ylim(0, 15) +   # Set the y-axis to go up to 15
  theme_minimal() +
  theme(axis.text.x = element_blank(),        # Remove x-axis text
        axis.ticks.x = element_blank(),       # Remove x-axis ticks
        legend.position = "bottom")

print(top10genera)
```


#new and old wgs classifications for ancestry analyses
```{r}
#adding the new WGS columns into the metadata 

library(readxl)

# Read the entire sheet
data <- read_excel("/Users/russ/Desktop/Aim 1+2/aims 1+2 seq data/metadata.xlsx")

data <- data[!is.na(data$seq_file), ]

# Extract just one column by name
new_wgs <- data$new_wgs
three_wgs <- data$three_wgs

#add each new column
metadata$new_wgs <- new_wgs
metadata$three_wgs <- three_wgs


#alpha comparisons of new wgs groupings 


shannon_diversity<-estimate_richness(ps_nochloro, measures = "shannon")
inverse_simpson_index <- estimate_richness(ps_nochloro, measures = "InvSimpson")
chao1<- estimate_richness(ps_nochloro, measures = "Chao1")
richness<- estimate_richness(ps_nochloro, measures = "Observed")

# Calculate diversity measures
diversity_df <- data.frame(
  SampleID = sample_names(ps_nochloro),
  Shannon = estimate_richness(ps_nochloro, measures = "Shannon")[, "Shannon"],
  Chao = estimate_richness(ps_nochloro, measures = "Chao1")[, "Chao1"],
  Richness = estimate_richness(ps_nochloro, measures = "Observed")[, "Observed"],
  OTUs = otu_table(ps_nochloro) %>% rowSums()
)

#making the metadata object and confirming dataframes 
meta_data<-sample_data(ps_nochloro)
meta_data <- data.frame(as(meta_data, "matrix"))
diversity_df<-as.data.frame(diversity_df)

# Rename seq_file to SampleID in meta_data
colnames(meta_data)[colnames(meta_data) == "seq_file"] <- "SampleID"

# Combine with metadata
diversity_df <- merge(diversity_df, meta_data, by = "SampleID")

names(diversity_df)[names(diversity_df) == "X3_wgs"] <- "three_wgs"


# Filter out NAs
diversity_df <- diversity_df %>%
  filter(!is.na(wgs.class) & !is.na(collection_time) & !is.na(new_wgs) & !is.na(same_wgs) & !is.na(three_wgs))

# Set desired order for wgs.class and collection_time
desired_order_wgs <- c("BCCH", "HYCH", "ADV.CA", "CACH") 
diversity_df$wgs.class <- factor(diversity_df$wgs.class, levels = desired_order_wgs)
desired_order_collection_time <- c("initial", "self_feeding", "pre_surgery") 
diversity_df$collection_time <- factor(diversity_df$collection_time, levels = desired_order_collection_time)



#making sure the variables are seen as factors not characters 
diversity_df$wgs.class<-as.factor(diversity_df$wgs.class)
diversity_df$collection_time<-as.factor(diversity_df$collection_time)
diversity_df$Collection_year<-as.factor(diversity_df$Collection_year)
diversity_df$Sex<-as.factor(diversity_df$Sex)
diversity_df$Nest_Site<-as.factor(diversity_df$Nest_Site)
diversity_df$Nest<-as.factor(diversity_df$Nest)
diversity_df$new_wgs<-as.factor(diversity_df$new_wgs)
diversity_df$three_wgs<-as.factor(diversity_df$three_wgs)




#SHANNON DIVERSITY
#same
shan.old <- lmer(Shannon ~ wgs.class*collection_time*Sex + (1|Collection_year/Nest), data = diversity_df)
summary(shan.old)
#check the Anova 
Anova(shan.old)
#  model simplification
shan.old.ms <- update(shan.old, ~ . 
  - Sex:collection_time:wgs.class)
anova(shan.old.ms,shan.old)
Anova(shan.old.ms)
shan.old.ms.1 <- update(shan.old.ms, ~ . 
  - Sex:collection_time)
anova(shan.old.ms.1,shan.old.ms)
Anova(shan.old.ms.1)
shan.old.ms.2 <- update(shan.old.ms.1, ~ . 
  - Sex:wgs.class)
anova(shan.old.ms.1,shan.old.ms.2)
Anova(shan.old.ms.2)
shan.old.ms.3 <- update(shan.old.ms.2, ~ . 
  - collection_time:wgs.class)
anova(shan.old.ms.3,shan.old.ms.2)
Anova(shan.old.ms.3)




#new
shan.new <- lmer(Shannon ~ new_wgs*collection_time*Sex + (1|Collection_year/Nest), data = diversity_df)
summary(shan.new)
#check the Anova 
Anova(shan.new)
shan.new.ms <- update(shan.new, ~ . 
  - Sex:collection_time:new_wgs)
anova(shan.new.ms,shan.new)
Anova(shan.new.ms)
shan.new.ms.1 <- update(shan.new.ms, ~ . 
  - Sex:new_wgs)
anova(shan.new.ms.1,shan.new.ms)
Anova(shan.new.ms.1)
shan.new.ms.2 <- update(shan.new.ms.1, ~ . 
  - Sex:collection_time)
anova(shan.new.ms.1,shan.new.ms.2)
Anova(shan.new.ms.2)
shan.new.ms.3 <- update(shan.new.ms.2, ~ . 
  - new_wgs:collection_time)
anova(shan.new.ms.3,shan.new.ms.2)
Anova(shan.new.ms.3)





#three
shan.three <- lmer(Shannon ~ three_wgs*collection_time*Sex + (1|Collection_year/Nest), data = diversity_df)
summary(shan.three)
#check the Anova 
Anova(shan.three)
shan.three.ms <- update(shan.three, ~ . 
  - three_wgs:Sex:collection_time)
anova(shan.three.ms,shan.three)
Anova(shan.three.ms)
shan.three.ms.1 <- update(shan.three.ms, ~ . 
  - three_wgs:Sex)
anova(shan.three.ms.1,shan.three.ms)
Anova(shan.three.ms.1)
shan.three.ms.2 <- update(shan.three.ms.1, ~ . 
  - Sex:collection_time )
anova(shan.three.ms.1,shan.three.ms.2)
Anova(shan.three.ms.2)
shan.three.ms.3 <- update(shan.three.ms.2, ~ . 
  - three_wgs:collection_time )
anova(shan.three.ms.3,shan.three.ms.2)
Anova(shan.three.ms.3)





#CHAO DIVERSITY
#same
chao.old <- lmer(Chao ~ wgs.class*collection_time*Sex + (1|Collection_year/Nest), data = diversity_df)
summary(chao.old)
#check the Anova 
Anova(chao.old)



#new
chao.new <- lmer(Chao ~ new_wgs*collection_time*Sex + (1|Collection_year/Nest), data = diversity_df)
summary(chao.new)
#check the Anova 
Anova(chao.new)
chao.new.ms <- update(chao.new, ~ . 
  - new_wgs:Sex:collection_time)
anova(chao.new.ms,chao.new)
Anova(chao.new.ms)
chao.new.ms.1 <- update(chao.new.ms, ~ . 
  - Sex:collection_time)
anova(chao.new.ms,chao.new.ms.1)
Anova(chao.new.ms.1)
chao.new.ms.2 <- update(chao.new.ms.1, ~ . 
  - Sex:new_wgs)
anova(chao.new.ms.2,chao.new.ms.1)
Anova(chao.new.ms.2)
chao.new.ms.3 <- update(chao.new.ms.2, ~ . 
  - collection_time:new_wgs)
anova(chao.new.ms.2,chao.new.ms.3)
Anova(chao.new.ms.3)



#three
chao.three <- lmer(Chao ~ three_wgs*collection_time*Sex + (1|Collection_year/Nest), data = diversity_df)
summary(chao.three)
#check the Anova 
Anova(chao.three)
chao.three.ms <- update(chao.three, ~ . 
  - three_wgs:Sex:collection_time)
anova(chao.three.ms,chao.three)
Anova(chao.three.ms)
chao.three.ms.1 <- update(chao.three.ms, ~ . 
  - Sex:collection_time)
anova(chao.three.ms,chao.three.ms.1)
Anova(chao.three.ms.1)
chao.three.ms.2 <- update(chao.three.ms.1, ~ . 
  - Sex:three_wgs)
anova(chao.three.ms.2,chao.three.ms.1)
Anova(chao.three.ms.2)
chao.three.ms.3 <- update(chao.three.ms.2, ~ . 
  - collection_time:three_wgs)
anova(chao.three.ms.3,chao.three.ms.2)
Anova(chao.three.ms.3)


#NUMBER OF ASVs
#same
otu.old <- lmer(OTUs ~ same_wgs*collection_time*Sex + (1|Collection_year/Nest) + (1|Sex), data = diversity_df)
summary(otu.old)
#check the Anova 
Anova(otu.old)
otu.old.ms <- update(otu.old, ~ . 
  - same_wgs:Sex:collection_time)
anova(otu.old.ms,otu.old)
Anova(otu.old.ms)
otu.old.ms.1 <- update(otu.old.ms, ~ . 
  - Sex:collection_time)
anova(otu.old.ms,otu.old.ms.1)
Anova(otu.old.ms.1)
otu.old.ms.2 <- update(otu.old.ms, ~ . 
  - Sex:same_wgs)
anova(otu.old.ms.2,otu.old.ms.1)
Anova(otu.old.ms.2)
otu.old.ms.3 <- update(otu.old.ms.2, ~ . 
  - Sex:collection_time)
anova(otu.old.ms.3,otu.old.ms.2)
Anova(otu.old.ms.3)


#new
otu.new <- lmer(OTUs ~ new_wgs*collection_time*Sex + (1|Collection_year/Nest), data = diversity_df)
summary(otu.new)
#check the Anova 
Anova(otu.new)
otu.new.ms <- update(otu.new, ~ . 
  - new_wgs:Sex:collection_time)
anova(otu.new.ms,otu.new)
Anova(otu.new.ms)
otu.new.ms.1 <- update(otu.new.ms, ~ . 
  - Sex:collection_time)
anova(otu.new.ms,otu.new.ms.1)
Anova(otu.new.ms.1)
otu.new.ms.2 <- update(otu.new.ms.1, ~ . 
  - Sex:new_wgs)
anova(otu.new.ms.2,otu.new.ms.1)
Anova(otu.new.ms.2)


#three
otu.three <- lmer(OTUs ~ three_wgs*collection_time*Sex + (1|Collection_year/Nest), data = diversity_df)
summary(otu.three)
#check the Anova 
Anova(otu.three)
otu.three.ms <- update(otu.three, ~ . 
  - three_wgs:Sex:collection_time)
anova(otu.three.ms,otu.three)
Anova(otu.three.ms)
otu.three.ms.1 <- update(otu.three.ms, ~ . 
  - Sex:collection_time)
anova(otu.three.ms,otu.three.ms.1)
Anova(otu.three.ms.1)
otu.three.ms.2 <- update(otu.three.ms.1, ~ . 
  - Sex:three_wgs)
anova(otu.three.ms.2,otu.three.ms.1)
Anova(otu.three.ms.2)



#beta comparisons of new wgs groupings 


#PERMANOVA of beta diversity ordination plot 
library(vegan)

#set meta_data
meta_data<-sample_data(ps_nochloro)
meta_data <- data.frame(as(meta_data, "matrix"))

names(meta_data)[names(meta_data) == "X3_wgs"] <- "three_wgs"

# Bray-Curtis dissimilarity
bray <- phyloseq::distance(ps_nochloro, method = "bray")

# Jaccard dissimilarity
jaccard <- phyloseq::distance(ps_nochloro, method = "jaccard")


#perform PERMANOVA
permanova_result_old <- adonis2(jaccard  ~ wgs.class+Sex+collection_time, data = metadata, permutations = 10000, na.action = "na.exclude", by = "margin")
print(permanova_result_old)

permanova_result_new <- adonis2(jaccard ~ new_wgs+Sex+collection_time, data = metadata, permutations = 10000, na.action = "na.exclude", by = "margin")
print(permanova_result_new)

permanova_result_three <- adonis2(jaccard ~ three_wgs+Sex+collection_time, data = metadata, permutations = 10000, na.action = "na.exclude", by = "margin")
print(permanova_result_three)



#remaking PCoA

# Load necessary libraries
library(phyloseq)
library(ggplot2)

# Create distance matrix (Bray-Curtis)
dist_matrix <- phyloseq::distance(ps_nochloro, method = "bray")

# Perform PCoA
pcoa_result <- ordinate(ps_nochloro, method = "PCoA", distance = dist_matrix)


point_colors <- c("initial" = "", "self_feeding" = "", "pre_surgery" = "")
fill_colors <- c("BCCH" = "aquamarine3", "HYCH" = "chocolate1", "ADV.CA" = "cornflowerblue", "CACH" = "pink3")

# Create the PCoA plot
pcoa_plot <- plot_ordination(ps_nochloro, ordination, type = "samples") +
  geom_point(aes(shape = collection_time, color = Sex), size = 3) +  
  stat_ellipse(aes(group = Sex, fill = Sex, color = Sex),  # Add color to outline
               geom = "polygon", alpha = 0.2, size = 1) +  # Increase line size
  # scale_fill_manual(values = fill_colors) +
  # scale_color_manual(values = fill_colors) +
  theme_minimal() +
  labs(title = "PCoA of Microbial Communities",
       shape = "Collection Time", color = "Collection Time")


# Display the plot
print(pcoa_plot)


pdf(paste("/Users/russ/Desktop/pcoa_plot.pdf"), width = 10, height = 10)
print(pcoa_plot)
dev.off()



#post hoc alpha - across new ancestry classifications 


library(emmeans)
# checking post hoc after the linear mixed effects above 

emmeans_results <- emmeans(otu.new.ms.2, specs = "new_wgs", by = "collection_time")
# Get the estimated marginal means for the main effects and interactions
#emmeans_results <- emmeans(otu.three.ms.2, pairwise ~ collection_time*three_wgs)

# Perform pairwise comparisons (adjust method can be changed as needed)
contrast_results <- pairs(emmeans_results, adjust = "tukey")

# Print the results
summary(contrast_results)




# Visualize interaction while accounting for nesting
shan_emm<-emmip(otu.three.ms.2, ~ collection_time | three_wgs)
print(shan_emm)

# Save the plot to a PDF
pdf("/Users/russ/Desktop/shan_emm.pdf", width = 10, height = 5)
print(shan_emm)
dev.off()
print(shan_emm)



#post hoc for the beta diversity testing 

library(pairwiseAdonis)

# Bray-Curtis dissimilarity
bray <- phyloseq::distance(ps_nochloro, method = "bray")

# Jaccard dissimilarity
jaccard <- phyloseq::distance(ps_nochloro, method = "jaccard")

meta_data$three_wgs[meta_data$three_wgs == "N/A"] <- NA
meta_data$new_wgs[meta_data$new_wgs == "N/A"] <- NA
metadata$wgs.class[metadata$wgs.class == "N/A"] <- NA



#pairwise adonis
pairwise_result <- pairwise.adonis2(jaccard ~ wgs.class, data = meta_data, permutations = 9999, na.action = "na.exclude")
print(pairwise_result)

#the betadisper and tukey of betadisper shows how different the variance within groups is 
#Tukey HSD on betadisper: Tests group difference in within-group variability.
#If Tukey HSD is not significant → group dispersions are the same → PERMANOVA is likely reflecting real differences in composition.

# running the post hoc tests 
dispersion_test <- betadisper(bray, meta_data$wgs.class)
print(dispersion_test)

dispersion_result <- anova(dispersion_test)
print(dispersion_result)

pairwise_dispersion <- TukeyHSD(dispersion_test)
print(pairwise_dispersion)




#looking for taxa related differences 

#taxa relation to grouping differences in composition differences 
# Remove rows with NA values in the 'new_wgs' column
meta_data_clean <- sample_data(ps_nochloro)[!is.na(sample_data(ps_nochloro)$new_wgs), ]
# Subset OTU table based on the cleaned metadata
ps_clean <- prune_samples(rownames(meta_data_clean), ps_nochloro)
# Example: Run simper with the cleaned phyloseq object
otu_table_clean <- otu_table(ps_clean)
simper_result <- simper(otu_table_clean, group = sample_data(ps_clean)$new_wgs)
summary(simper_result)



#line graph of hybrid index for each of the diff scales 

library(ggplot2)

meta_data$hybrid.index <- as.numeric(meta_data$hybrid.index)


library(dplyr)

# Filter to unique band_id values
meta_data_unique <- meta_data %>% 
  distinct(band_id, .keep_all = TRUE)



new_wgs_line<-ggplot(meta_data_unique, aes(x = hybrid.index, y = 1, 
                      color = new_wgs, shape = new_wgs)) +
  geom_jitter(height = 0.05, width = 0, size = 3) +
  scale_x_continuous(limits = c(0, 1)) +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(title = "Samples Along the Hybrid Index",
       x = "Hybrid Index (0–1)",
       color = "WGS Class",
       shape = "WGS Class")


print(new_wgs_line)

pdf("/Users/russ/Desktop/new_wgs_line.pdf", width = 5, height = 5)
print(new_wgs_line)
dev.off()


old_wgs_line<-ggplot(meta_data_unique, aes(x = hybrid.index, y = 1, 
                      color = same_wgs, shape = same_wgs)) +
  geom_jitter(height = 0.05, width = 0, size = 3) +
  scale_x_continuous(limits = c(0, 1)) +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(title = "Samples Along the Hybrid Index",
       x = "Hybrid Index (0–1)",
       color = "WGS Class",
       shape = "WGS Class")

print(old_wgs_line)

pdf("/Users/russ/Desktop/old_wgs_line.pdf", width = 5, height = 5)
print(old_wgs_line)
dev.off()


three_wgs_line<-ggplot(meta_data_unique, aes(x = hybrid.index, y = 1, 
                      color = three_wgs, shape = three_wgs)) +
  geom_jitter(height = 0.05, width = 0, size = 3) +
  scale_x_continuous(limits = c(0, 1)) +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank()) +
  labs(title = "Samples Along the Hybrid Index",
       x = "Hybrid Index (0–1)",
       color = "WGS Class",
       shape = "WGS Class")

print(three_wgs_line)

pdf("/Users/russ/Desktop/three_wgs_line.pdf", width = 5, height = 5)
print(three_wgs_line)
dev.off()

```



